Parallelizing Linear Recurrent Neural Nets Over Sequence Length

### 2017-10-27 paper/main.tex
Paper edits etc

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{amsmath}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{graphicx}

@@

@@

\begin{abstract}
Recurrent neural networks (RNNs) are widely used to model sequential data but
their non-linear dependencies between sequence elements prevent parallelizing
training over sequence length. We show the training of RNNs with only linear
sequential dependencies can be parallelized over the sequence length using the
parallel scan algorithm, leading to rapid training on long sequences with small
minibatch size. We develop a parallel linear recurrence CUDA kernel and show
that it can be applied to immediately speed up several RNN architectures by up
to 9x. We abstract prior linear sequence models into a new framework of linear
surrogate RNNs and develop a linear surrogate for the long short-term memory 
unit, the LS-LSTM, which is able to achieve throughput 30x greater than the
currently fastest-available LSTM implementation. We show that the LS-LSTM is
able to perform well in the 
\end{abstract}

\section{Introduction}

Recurrent neural networks (RNNs) are widely used for sequence modelling tasks in domains such as 
natural language processing \cite{sutskever2014sequence}, 
speech recognition \cite{amodei2015deep},
and reinforcement learning \cite{hausknecht2015deep}. Most RNNs, including popular variants such as long short-term memories (LSTMs) \cite{hochreiter1997long} and gated recurrent units (GRUs) \cite{cho2014learning}, contain a non-linear dependency
between sequential inputs. These non-linear dependencies create a very flexible class of
models but limit the feasibility of training RNNs on long sequences as each sequence element
must be processed sequentially.
Modelling sequences of thousands to millions of elements is important to domains
such as robotics, remote sensing, control systems, speech recognition, medicine, and finance.

The RNN serial evaluation inefficiency problem is usually mitigated by parallelizing the forward
and backward pass over a minibatch of inputs. Without minibatches, RNN evaluation is a sequence of matrix-vector multiplications. Minibatches transform RNN computation into a sequence of more efficient matrix-matrix multiplications, but minibatches within RNNs present many issues.
RNN model size is often limited by GPU memory size, and running a forward 
and backward pass on a minibatch requires memory linear in
the minibatch size.
Grouping data into minibatches increases the latency of each pass and reduces the rate of optimization steps. Finally, training with larger minibatches damages generalization ability 
\cite{keskar2017large}. Persistent RNNs \cite{diamos2016persistent} use a novel implementation that can achieve high GPU utilization with very small 
minibatch sizes when the recurrent state is larger than 500 elements, but even
persistent RNNs become limited by the serial evaluation
inefficiency at smaller hidden sizes.

Numerous prior works have shown strong performance from neural sequential models with only linear dependence on earlier sequence elements. \citet{balduzzi2016strongly} investigated RNNs with only elementwise linear recurrence relations 
$h_t = \alpha_t \odot h_{t-1} + (1-\alpha_t) \odot x_t$ and developed
linear variants of LSTM and GRU that perform similarly to
standard non-linear RNNs on text generation tasks. \citet{bradbury2017quasi},  \citet{kalchbrenner2016neural}, \citet{gehring2017convolutional}, and \citet{van2016wavenet} have successfully applied networks of convolutions over sequences for tasks
such as machine translation, language modelling, and audio generation. 
These works have observed up to an order of magnitude 
increase in training throughput compared to RNN alternatives. Convolutional sequence models typically rely on either an attention mechanism or a (possibly linear) recurrent layer
to integrate information at scales larger than the filter width. Introduction of a recurrent
layer prevents full parallelization over the sequence length while attention mechanisms are
expensive to apply on long sequences in online inference use cases.
One dimensional convolution can be viewed as a learnable linear
finite impulse response (FIR) filter with a parallel evaluation algorithm, while
linear recurrence is a learnable linear infinite impulse response (IIR). This work
parallelizes evaluation of linear recurrences through application of the parallel
scan algorithm.

@@

Our contributions are as follows:
\begin{itemize}
\item{We describe the class of RNNs which satisfy the conditions above, and
    show that many of the RNNs used in practice such as the Quasi-RNN are
    contained in this class}
\item{We provide an implementation of the parallel linear recurrence
    algorithm as a CUDA kernel, and show that it speeds up training of
    the SRU and QRNN architectures by factors of up to 9x}
\item{We describe how several recent linear RNNs can be described as
    linear surrogates for nonlinear architectures. We introduce a linear
    surrogate for the LSTM and show that we are able to train it
    with a speedup of 5-10x compared to the CuDNN-LSTM when we use the plr
    algorithm.}
\end{itemize}

\section{Parallel linear recurrence}
Considering the familiar operations in linear algebra, we see that the
associative operation of vector addition ($\boldsymbol{x} \oplus \boldsymbol{y}
= \boldsymbol{x} + \boldsymbol{y}$), the semiassociative operation of
matrix-vector multiplication ($A \otimes \boldsymbol{x} = A\boldsymbol{x}$) and
the associative operation of matrix-matrix multiplication ($A \odot B=AB$) form
recurrences which satisfy Blelloch's three conditions, thus allowing the linear
recurrence $h_t = \Lambda_t h_{t-1} + x_t$ to be evaluated in parallel over $t$
for vectors $x_t$ and square matrices $\Lambda_t$. As the method is essential to
this work, Algorithm 1 presents the parallel linear recurrence algorithm for the
interested reader.
\begin{algorithm}
  \label{alg:plr}
\caption{Parallel linear recurrence on $p$ processors}
\begin{algorithmic}[1]
  \State Let $y = [(\Lambda_1, x_1), (\Lambda_2, x_2), ..., (\Lambda_T, x_T)]$
  \State Let binary operator $\bullet$ act as $(\Lambda, x) \bullet h = \Lambda h + x$
  \State Let $S_0=1, S_i < E_i, E_i + 1 = S_{i+1}, E_{p-1}=T$ for $i$ in $0,p-1$

@@

\subsection{Theoretical performance}
The cost of a serial scan over a sequence of length $T$ is 
$C_\text{sscan} = (C_\otimes + C_\oplus)T$, compared to the parallel scan cost
$C_\text{pscan} = 2(C_\odot + C_\otimes + C_\oplus)(T/p + \lg p)$ \cite{blelloch1990prefix},
where \(p\) is the number of available parallel cores. 
If $h_t$ is a vector of dimension $n$ then 
$C_\odot\propto n^3, C_\otimes\propto n^2, C_\oplus\propto n$ giving
$C_\text{pscan} \propto 2(n^3 + n^2 + n)(T/p + \lg p)$ and 
$C_\text{sscan} \propto (n^2 + n)T$. The $n^3$ cost of the matrix
multiplication in the parallel algorithm can counter-act any parallel speedups for
sufficiently large hidden states and lead to a slower algorithm overall.

To avoid this problem, we will only consider diagonal matrices $\Lambda_t$, in
which case both matrix-matrix and matrix-vector multiplication have cost
proportional to $n$ and $C_\text{pscan}\propto 6n(T/p + \lg p)$ and
$C_\text{sscan}=2nT$. Assuming $p \ll T$, then $C_\text{pscan} \le
C_\text{sscan}$ when $p \ge 3$.\footnote{Can include the actual speedup
  factor, \(\alpha = \frac{PT}{3(T + \lg(p))}\).}
As we are only considering diagonal matrices,
we write the linear recurrence as $h_t = \lambda_t \odot h_{t-1} + x_t$
where $\odot$ indicates elementwise multiplication.

Limiting $\Lambda_t$ to be diagonal may seem like a severe constraint but there are
several reasons to do so beyond the unfavorable parallelization performance. Relatively few neural
network models use separate recurrent matrices for each sequence element and using these
separate matrices would require potentially prohibitive $n^2T$ memory.\footnote{Ideally we can
  find a citation for someone thinking these points before}
Applying
the same matrix $\Lambda$ to each sequence element is also unappealing considering that a matrix
multiplication can be thought of as a rotation and a scaling. The same rotation at every
element seems unlikely to be useful, and the scaling is exactly what's captured in diagonal
vectors $\lambda_t$. Recurrent coefficient vectors $\lambda_t$ provide enough flexibility
to implement schemes such as exponential moving averages or a gating mechanism.

@@

\subsection{Gated impulse linear recurrent layer}
A gated impulse linear recurrent (GILR) layer transforms its $m$ dimensional inputs $x_t$ into a sequence of $n$ dimensional hidden states $h_t$:
\begin{align*}
g_t &= \sigma(Ux_t + b_g) \\
i_t &= \tau(Vx_t + b_z) \\
h_t &= g_t \odot h_{t-1} + (1-g_t)\odot i_t
\end{align*}
A GILR layer applies the same non-linear transform to each sequence element and then accumulates the sequence elements with a non-linear gating mechanism. Gate $g_t$ uses the sigmoid activation function to have values in [0,1] for reasonable gating semantics, while impulse $i_t$ can use any activation function $\tau$. Stacking GILR layers allows for rich non-linear dependence on previous events while still taking advantage of fast parallel sequence evaluation.

@@

Now consider the GILR, noting that it has the same two matrix-vector
multiplications per iteration as the above RNN. $g$ and $i$ can each be
evaluated for all $t$ with a single $(bT, m) \times (m, n)$ matrix
multiplication each. Given $g$ and $i$, $h$ can be computed using a parallel
linear recurrence over $T$ vectors each of $bn$ elements. Rather than $T$ small
operations, the GILR can be evaluated over all sequence elements with two large
matrix multiplies and a parallel linear recurrence. GILR performance is much
less dependent on batch size as the matrix multiplies see an "effective batch
size" of $bT$ and $T$ is typically large.

\subsection{Linear surrogate RNNs}
\label{sec:ls-rnns}
RNNs learn a transition function $s_t = f(s_{t-1}, x_t)$ which combines previous
state $s_{t-1}$ with input $x_t$ to compute current state $s_t$. Non-linear $f$
prevents application of the parallel linear recurrence algorithm and forces slow
serial evaluation. To work around this inefficiency, note that $s_t$ serves a
dual purpose. In $s_t = f(s_{t-1}, x_t)$, $s_{t-1}$ serves as an input to $f$
summarizing the previous inputs while $s_t$ serves as the output of $f$ to be
passed to other layers of the network. If we decouple these uses, we can instead
compute $s_t = f(\tilde{s}_{t-1}, x_t)$ with $\tilde{s}_t$ as a linearly
computable surrogate for $s_t$. With this linear surrogate, non-linear $f$ can
still be evaluated. We refer to this class of model as a linear surrogate RNN
(LS-RNN). Quasi-RNNs \cite{bradbury2017quasi} are LS-RNNs using $\tilde{h}_{t-1}
= W_k x_{t-k} + ... W_1 x_{t-1}$ and strongly typed
RNNs\cite{balduzzi2016strongly} are LS-RNNs with $\tilde{h}_t=x_{t-1}$. Although
not a rule, LS-RNNs can often be parallelized over sequence length with either
convolution or linear recurrence.

As an example LS-RNN, consider an LSTM\footnote{Want a citation here I guess}:
\begin{align*}
f_t, i_t, o_t &= \sigma(U_{f,i,o} h_{t-1} + V_{f,i,o} x_t + b_{f,i,o}) \\
z_t &= \tau(U_z h_{t-1} + V_z x_t + b_z) \\
c_t &= f_t \odot c_{t-1} + i_t \odot z_t \\
h_t &= o_t \odot c_t
\end{align*}
An LSTM has state $s_t = (h_t, c_t)$. Since $c_t$ depends only linearly on $c_{t-1}$, no surrogate is needed for $c_t$. $h_t$ has a non-linear dependence on $h_{t-1}$, so $h_t$ needs a linear surrogate. With a GILR layer as surrogate, the linear surrogate LSTM (LS-LSTM) is
\begin{align*}
f_t, i_t, o_t &= \sigma(U_{f,i,o} \tilde{h}_{t-1} + V_{f,i,o} x_t + b_{f,i,o}) \\
z_t &= \tau(U_z \tilde{h}_{t-1} + V_z x_t + b_z) \\
c_t &= f_t \odot c_{t-1} + i_t \odot z_t \\
h_t &= o_t \odot c_t \\
g_t &= \sigma(V_g x_t + b_g) \\
\tilde{h}_t &= g_t \odot \tilde{h}_{t-1} + (1-g_t)\odot \tau(Wx_t + b_h)
\end{align*}

For $m$ inputs and hidden size $n$, the LS-LSTM contains $2n(n+m)$ more parameters than the equivalently sized LSTM to handle the mapping from $x$ to $\tilde{h}$. More generally, a LS-RNN contains all of the same parameters as the underlying RNN as well as some additional parameters to compute the linear surrogate.

\section{Experiments}
We perform several experiments, first confirming that our implementation of the
parallel linear recurrence algorithm is able to achieve higher throughput than a
similarly implemented serial version of the algorithm. We then show that the
LS-LSTM architecture, trained with the PLR algorithm, is able to outperform the
optimized CuDNN LSTM implementation on a pathological example from the original
LSTM paper. Finally we show that our LS-LSTM is able to quickly achieve state of
the art results on an existing medical data set.

\subsection{Throughput Benchmarks}
**** Throughput of pure PLR kernel vs SLR kernel
**** Show how it speeds up the SRU, QRNN, and LS-LSTM
**** Show how the LS-LSTM has much better throughput than the CudnnLSTM

We first illustrate the throughput advantage of the parallel scan algorithm for
evaluating the linear recurrence. We define throughput as events/s and a
minibatch of size $b$ and length $T$ to be $bT$ events. We implement two CUDA
kernels, one which evaluates the parallel linear recurrence described in
algorithm \ref{alg:plr}, and one which evaluates the linear recurrence using the
straightforward serial approach. This comparison is performed directly at the
kernel level, avoiding any overhead due to calling Tensorflow. We find that at
long sequence lengths, the PLR kernel has a distinct
advantage, as shown in table \ref{table:kernel-throughput}.

\begin{figure}
\begin{center}
  \begin{tabular}{lrrr}
    \label{table:rnn-throughput}
Sequence Length\footnote{What's the batch size etc here??} & 4 Dimensions & 32 Dimensions & 128 Dimensions \\
\hline
16 & 0.06 & 0.06 & 0.05\\ 
256 & 0.22 & 0.22 & 0.86\\ 
4,096 & 1.02 & 2.94 & 3.36\\ 
65,536 & 38.5 & 41.8 & 17.5\\
  \end{tabular}  
    \end{center}
    \caption{Speedup of the PLR kernel compared to the SLR kernel, without any
    overhead from Tensorflow.}    
  \end{figure}

Many RNN architectures that have been recently introduced are amenable to
implementing through parallel linear recurrence. Table
\ref{table:rnn-throughput} shows the throughput advantage by using a parallel
linear recurrence when compared to a serial linear recurrence. Since PLR and SLR
are simply different methods for computing the same output, there is no penalty
incurred for this speedup: it is a pure speed gain. Typically experiments will aim
to maximize the memory utilisation of the GPU, so we simulate this by choosing the
product of the batch size and the sequence length to be constant. We see that the
architectures which have fewer terms (for which the linear recurrence is a higher
proportion of the computational burden) are more effected by the switch to PLR.
This is particularly clear in the case of the QRNN, where including a wider
convolutional window results in less time spent on the recurrence and less to gain
from the parallelization. 
\begin{figure}
\begin{center}
  \begin{tabular}{lrrrr}
    \label{table:rnn-throughput}
Sequence Length & LS-LSTM & SRU & QRNN(2) & QRNN(10)\\
\hline
16 & 0.61 & 0.28 & 0.38 & 0.78\\
256 & 0.91 & 0.84 & 0.86 & 0.99\\
4,096 & 0.98 & 1.38 & 1.18 & 1.05\\
65,536 & 1.41 & 9.21 & 6.68 & 2.05\\
  \end{tabular}  
\end{center}
\caption{Speedup obtained when using parallel linear recurrence compared
    to serial linear recurrence. We analyse this for several recent architectures:
    the Simple Recurrent Unit from [cite], the Quasi-RNN from [cite], with a
    convolutional filter of size 2 and 10, and the Linear Surrogate LSTM that
    we introduce in section \ref{sec:ls-rnns}. To give a fair model of a
    typical use-case, we set batch size \(b\) such that \(Tb = 65536\)
    for a sequence length \(T\).}
\end{figure}

Finally, we compare the throughput of the linear surrogate LSTM to the current
fastest available implementation of the LSTM, the CuDNN-LSTM. We observe that our
LS-LSTM has higher throughput for most of the low-batch-size regime, with a 30x
speedup at a sequence length of 65,536. Recent work (cite https://arxiv.org/pdf/1609.04836.pdf)
has suggested that small mini-batches may aid generalization by encouraging
the discovery of shallow minima. The LS-LSTM allows us to achieve high throughput
for all choices of minibatch size and sequence length.

>>>>

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{amsmath}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{graphicx}
\usepackage{todonotes}
\usepackage{xargs}
\usepackage{booktabs}

@@

\newcommandx{\Chris}[2][1=]{\todo[linecolor=red,backgroundcolor=red!25,bordercolor=red,#1]{#2}}
\newcommandx{\Eric}[2][1=]{\todo[linecolor=blue,backgroundcolor=blue!25,bordercolor=blue,#1]{#2}}
\newcommandx{\thiswillnotshow}[2][1=]{\todo[disable,#1]{#2}}

@@

\begin{abstract}
Recurrent neural networks (RNNs) are widely used to model sequential data but
their non-linear dependencies between sequence elements prevent parallelizing
training over sequence length. We show the training of RNNs with only linear
sequential dependencies can be parallelized over the sequence length using the
parallel scan algorithm, leading to rapid training on long sequences with small
minibatch size.\Chris{Thought: we could rephrase this as allowing us to break
the necessity of using large batches to train fast} We develop a
parallel linear recurrence CUDA kernel and show that it can be applied to
immediately speed up several RNN architectures by up to a factor of 9. We abstract
recent work on linear RNNs into a new framework of linear surrogate RNNs and develop
a linear surrogate for the long short-term memory unit, the GILR-LSTM, which is
able to achieve throughput 30 times greater than the currently fastest-available LSTM
implementation. We show that the GILR-LSTM is able to perform well on real-world
problems, 
\end{abstract}

\section{Introduction}
Recurrent neural networks (RNNs) are widely used for sequence modelling tasks in
domains such as natural language processing \cite{sutskever2014sequence}, speech
recognition \cite{amodei2015deep}, and reinforcement learning
\cite{hausknecht2015deep}. Most RNNs, including popular variants such as long
short-term memories (LSTMs) \cite{hochreiter1997long} and gated recurrent units
(GRUs) \cite{cho2014learning}, contain a non-linear dependency between
sequential inputs. These non-linear dependencies create a very flexible class of
models but limit the feasibility of training RNNs on long sequences as each
sequence element must be processed sequentially.  Modelling sequences of
thousands to millions of elements is important to domains such as robotics,
remote sensing, control systems, speech recognition, medicine, and finance.

The RNN serial evaluation inefficiency problem is usually mitigated by
parallelizing the forward and backward pass over a minibatch of inputs. Without
minibatches, RNN evaluation is a sequence of matrix-vector
multiplications. Minibatches transform RNN computation into a sequence of more
efficient matrix-matrix multiplications, but this speed-up brings several
disadvantages.  RNN model size is often limited by GPU memory size, and running
a forward and backward pass on a minibatch requires memory linear in the
minibatch size.  Grouping data into minibatches increases the latency of each
pass and reduces the rate of optimization steps. Finally, training with larger
minibatches damages generalization ability \cite{keskar2017large}. It is clearly
important to investigate if it is possible to parallelise RNN training without
introducing minibatches. Persistent RNNs \cite{diamos2016persistent} use a novel
implementation that can achieve high GPU utilization with very small minibatch
sizes when the recurrent state is larger than 500 elements, but even persistent
RNNs become limited by the serial evaluation inefficiency at smaller hidden
sizes.

Numerous prior works have shown strong performance from neural sequential models
with only linear dependence on earlier sequence
elements. \citet{balduzzi2016strongly} investigated RNNs with only elementwise
linear recurrence relations $h_t = \alpha_t \odot h_{t-1} + (1-\alpha_t) \odot
x_t$ and developed linear variants of LSTM and GRU that perform similarly to
standard non-linear RNNs on text generation tasks. \citet{bradbury2017quasi},
\citet{kalchbrenner2016neural}, \citet{gehring2017convolutional}, and
\citet{van2016wavenet} have successfully applied networks of convolutions over
sequences for tasks such as machine translation, language modelling, and audio
generation.  These works have observed up to an order of magnitude increase in
training throughput compared to RNN alternatives. Convolutional sequence models
typically rely on either an attention mechanism or a (possibly linear) recurrent
layer to integrate information at scales larger than the filter
width. Introduction of a recurrent layer prevents full parallelization over the
sequence length while attention mechanisms are expensive to apply on long
sequences in online inference use cases.  One dimensional convolution can be
viewed as a learnable linear finite impulse response (FIR) filter with a
parallel evaluation algorithm, while linear recurrence is a learnable linear
infinite impulse response (IIR).\Chris{Seems like an aside - take out?}

  This work parallelizes evaluation of linear recurrences through application of
the parallel scan algorithm.

@@

Considering the familiar operations in linear algebra, we see that the
associative operation of vector addition
($\boldsymbol{x} \oplus \boldsymbol{y} = \boldsymbol{x} + \boldsymbol{y}$), the
semiassociative operation of matrix-vector multiplication
($A \otimes \boldsymbol{x} = A\boldsymbol{x}$) and the associative operation of
matrix-matrix multiplication ($A \odot B=AB$) satisfy Blelloch's three
conditions, allowing $h_t = \Lambda_t h_{t-1} + x_t$ to be evaluated in parallel
over time steps \(t\) for vectors $x_t$ and square matrices $\Lambda_t$.

We investigate this idea further and deliver the following contributions:
\begin{itemize}
\item{We classify the RNNs which satisfy the conditions above, and
    show that many of the RNNs used in practice such as the Quasi-RNN are
    contained in this class}
\item{We provide an implementation of the parallel linear recurrence
    algorithm as a CUDA kernel, and show that it speeds up training of
    the SRU and QRNN architectures by factors of up to 9x}
\item{We describe how several recent linear RNNs can be described as
    linear surrogates for nonlinear architectures. We introduce a linear
    surrogate for the LSTM and show that we are able to train it
    with a speedup of 5-10x compared to the CuDNN-LSTM when we use the
    parallel linear recurrence algorithm.}
\end{itemize}

\section{Parallel linear recurrence}
As the method is
essential to this work, Algorithm 1 presents the parallel linear recurrence
algorithm for the interested reader.
\begin{algorithm}
  \label{alg:plr}
\caption{Parallel linear recurrence on $p$ processors}
\begin{algorithmic}[1]
  \State Let $y = [(\Lambda_1, x_1), (\Lambda_2, x_2), ..., (\Lambda_T, x_T)]$
  \State Let binary operator $\bullet$ act as $(\Lambda, x) \bullet h = \Lambda h + x$
  \State Let $S_0=1, S_i < E_i, E_i + 1 = S_{i+1}, E_{p-1}=T$ for $i$ in $0,p-1$

@@

\subsection{Theoretical performance}
The cost of a serial scan over a sequence of length $T$ is 
$C_\text{sscan} \in \mathcal{O}((C_\otimes + C_\oplus)T)$, compared to the parallel scan cost
$C_\text{pscan} \in \mathcal{O}(2(C_\odot + C_\otimes + C_\oplus)(T/p + \lg p))$ \cite{blelloch1990prefix},
where \(p\) is the number of available parallel cores. 
If $h_t$ is a vector of dimension $n$ then 
$C_\odot \in  \mathcal{O}(n^3), C_\otimes \in \mathcal{O}(n^2), C_\oplus \in \mathcal{O}(n)$ giving
$C_\text{pscan} \in \mathcal{O}(2(n^3 + n^2 + n)(T/p + \lg p))$ and 
$C_\text{sscan} \in \mathcal{O}((n^2 + n)T)$. The $\mathcal{O}(n^3)$ cost of the matrix
multiplication in the parallel algorithm can counter-act any parallel speedups for
sufficiently large hidden states and lead to a slower algorithm overall.

To avoid this problem, we will only consider diagonal matrices $\Lambda_t$, in
which case both matrix-matrix and matrix-vector multiplication have cost
proportional to $n$ and $C_\text{pscan}\in \mathcal{O}(6n(T/p + \lg p))$ and
$C_\text{sscan} \in \mathcal{O}(2nT)$. Assuming $p \ll T$, then $C_\text{pscan} \le
C_\text{sscan}$ when $p \ge 3$.\Chris{Can include the actual speedup
  factor, \(\alpha = \frac{PT}{3(T + \lg(p))}\).}
As we are only considering diagonal matrices,
we write the linear recurrence as $h_t = \lambda_t \odot h_{t-1} + x_t$
where $\odot$ indicates elementwise multiplication.

Limiting $\Lambda_t$ to be diagonal may seem like a severe constraint but there are
several reasons to do so beyond the unfavorable parallelization performance. Relatively few neural
network models use separate recurrent matrices for each sequence element and using these
separate matrices would require potentially prohibitive $n^2T$ memory.\Chris{I can't
seem to find anyone explicitly saying this before from a quick google.}
Applying
the same matrix $\Lambda$ to each sequence element is also unappealing considering that a matrix
multiplication can be thought of as a rotation and a scaling. The same rotation at every
element seems unlikely to be useful, and the scaling is exactly what's captured in diagonal
vectors $\lambda_t$. Recurrent coefficient vectors $\lambda_t$ provide enough flexibility
to implement schemes such as exponential moving averages or a gating mechanism.

@@

\subsection{Gated impulse linear recurrent layer}
A gated impulse linear recurrent (GILR) layer transforms its $m$ dimensional inputs $x_t$ into a sequence of $n$ dimensional hidden states $h_t$:
\begin{align*}
g_t &= \sigma(Ux_t + b_g) \\
i_t &= \tau(Vx_t + b_z) \\
h_t &= g_t \odot h_{t-1} + (1-g_t)\odot i_t
\end{align*}
A GILR layer applies the same non-linear transform to each sequence element and
then accumulates the sequence elements with a non-linear gating mechanism. Gate
$g_t$ uses the sigmoid activation function to give values in [0,1] for
reasonable gating semantics, while impulse $i_t$ can use any activation function
$\tau$. Stacking GILR layers allows for rich non-linear dependence on previous
events while still taking advantage of fast parallel sequence evaluation.

@@

Now consider the GILR, noting that it has the same two matrix-vector
multiplications per iteration as the above RNN. The intermediate variables $g$
and $i$ can be evaluated for all $t$ with a single $(bT, m) \times (m, n)$
matrix multiplication each. Given $g$ and $i$, $h$ can be computed using a
parallel linear recurrence over $T$ vectors each of $bn$ elements. Rather than
$T$ small operations, the GILR can be evaluated over all sequence elements with
two large matrix multiplications and a parallel linear recurrence. GILR performance
is much less dependent on batch size as the matrix multiplies see an "effective
batch size" of $bT$ and $T$ is typically large.

\subsection{Linear surrogate RNNs}
\label{sec:ls-rnns}
RNNs learn a transition function $s_t = f(s_{t-1}, x_t)$ which combines previous
state $s_{t-1}$ with input $x_t$ to compute current state $s_t$. Non-linear $f$
prevents application of the parallel linear recurrence algorithm and forces slow
serial evaluation. To work around this inefficiency, note that $s_t$ serves a
dual purpose. In $s_t = f(s_{t-1}, x_t)$, $s_{t-1}$ serves as an input to $f$
summarizing the previous inputs while $s_t$ serves as the output of $f$ to be
passed to other layers of the network. We can decouple these uses and introduce
independent variables for each purpose: \(s_t\) is passed onto other layers of the network
and we introduce the linear surrogate \(\tilde{s}_t\) which is passed onto the
next state, with \(s_t = f(\tilde{s}_{t-1}, x_t)\). We are still able to choose a 
nonlinear \(f\), our only limitation being that \(\tilde{s}_t\) must be linearly
computable.  We refer to this class of model as a linear surrogate RNN
(LS-RNN). Quasi-RNNs \cite{bradbury2017quasi} are LS-RNNs using $\tilde{h}_{t-1}
= W_k x_{t-k} + ... W_1 x_{t-1}$ and strongly typed
RNNs\cite{balduzzi2016strongly} are LS-RNNs with $\tilde{h}_t=x_{t-1}$. Although
not a rule, LS-RNNs can often be parallelized over sequence length with either
convolution or linear recurrence.

As an example LS-RNN, consider an LSTM:
\begin{align*} f_t, i_t, o_t &= \sigma(U_{f,i,o} h_{t-1} + V_{f,i,o} x_t +
b_{f,i,o}) \\ z_t &= \tau(U_z h_{t-1} + V_z x_t + b_z) \\ c_t &= f_t \odot
c_{t-1} + i_t \odot z_t \\ h_t &= o_t \odot c_t
\end{align*} An LSTM has state $s_t = (h_t, c_t)$. Since $c_t$ depends only
linearly on $c_{t-1}$, no surrogate is needed for $c_t$. $h_t$ has a non-linear
dependence on $h_{t-1}$, so $h_t$ needs a linear surrogate. Introducing a GILR layer as
the surrogate, we obtain the GILR-LSTM:
\begin{align*} f_t, i_t, o_t &= \sigma(U_{f,i,o} \tilde{h}_{t-1} + V_{f,i,o} x_t
+ b_{f,i,o}) \\ z_t &= \tau(U_z \tilde{h}_{t-1} + V_z x_t + b_z) \\ c_t &= f_t
\odot c_{t-1} + i_t \odot z_t \\ h_t &= o_t \odot c_t \\ g_t &= \sigma(V_g x_t +
b_g) \\ \tilde{h}_t &= g_t \odot \tilde{h}_{t-1} + (1-g_t)\odot \tau(Wx_t + b_h)
\end{align*}

For $m$ inputs and hidden size $n$, the GILR-LSTM contains $2n(n+m)$ more
parameters than the equivalently sized LSTM to handle the mapping from $x$ to
$\tilde{h}$. More generally, a LS-RNN contains all of the same parameters as the
underlying RNN as well as some additional parameters to compute the linear
surrogate.

\section{Experiments}
We perform several experiments, first confirming that our implementation of the
parallel linear recurrence algorithm is able to achieve higher throughput than a
similarly implemented serial version of the algorithm. We show that the plr kernel
is up to 40 times faster than the serial equivalent for long sequence lengths, and
that this speedup translates to implementations of LS-RNNs such as the QRNN (where
we can get a speedup of up to 9 times).

In order to illustrate that the linearisation does not necessarily come at the
cost of expressibility, we show that the GILR-LSTM architecture, trained with
the PLR algorithm, is able to outperform the optimized CuDNN LSTM implementation
on a pathological example from the original LSTM paper
\cite{hochreiter1997long}. Finally we show that our GILR-LSTM is able to
quickly achieve state of the art results on an existing medical data
set. \Chris{I hope so\ldots}

\subsection{Throughput Benchmarks}
We first illustrate the throughput advantage of the parallel scan algorithm for
evaluating the linear recurrence. For a minibatch comprised of \(b\) sequences
of length \(T\), we define the number of events as \(bT\) and the throughput as
the number of events per second. We implement two CUDA kernels, one which
evaluates the parallel linear recurrence described in algorithm \ref{alg:plr},
and one which evaluates the linear recurrence using the straightforward serial
approach. This comparison is performed directly at the kernel level, avoiding
any overhead due to calling Tensorflow. We find that at long sequence lengths,
the PLR kernel has a distinct advantage, as shown in table
\ref{table:kernel-throughput}, with a speedup factor of 40. At short sequence
lengths, the parallel kernel does not perform well due to the extra overhead
involved in marshalling the parallel algorithm. \Chris{Important:
  What's the batch size here? Is it just one sequence? Need to say that}
\begin{table}[]
  \centering
\caption{Speedup of the PLR kernel compared to the SLR kernel, without any
  overhead from Tensorflow, for various sequence lengths and input size \(m\).}      
\label{table:kernel-throughput}
\begin{tabular}{@{}llll@{}}
Sequence Length & \(m=4\)  & \(m=32\) & \(m=128\) \\ \midrule
16              & 0.06 & 0.06 & 0.05  \\
256             & 0.22 & 0.22 & 0.86  \\
4,096           & 1.02 & 2.94 & 3.36  \\
65,536          & 38.5 & 41.8 & 17.5  \\ \bottomrule
\end{tabular}
\end{table}

Many of the RNNs that have been recently introduced can be implemented with a
parallel linear recurrence. Table \ref{table:rnn-throughput} shows the
throughput advantage by using a parallel linear recurrence when compared to a
serial linear recurrence. Since PLR and SLR are simply different methods for
computing the recurrence, there is no penalty incurred for this speedup - we
obtain exactly the same output.

Typically experiments will aim to maximize the memory
utilisation of the GPU, so we simulate this by choosing the product of the batch
size and the sequence length to be constant. We see that the architectures which
have fewer terms (for which the linear recurrence is a higher proportion of the
computational burden) are more effected by the switch to PLR.  This is
particularly clear in the case of the QRNN, where including a wider
convolutional window results in less time spent on the recurrence and less to
gain from the parallelization.
\begin{figure}
\begin{center}
  \begin{tabular}{@{}lrrrr@{}}
    \label{table:rnn-throughput}
Sequence Length & GILR-LSTM & SRU & QRNN(2) & QRNN(10)\\ \midrule
16 & 0.61 & 0.28 & 0.38 & 0.78\\
256 & 0.91 & 0.84 & 0.86 & 0.99\\
4,096 & 0.98 & 1.38 & 1.18 & 1.05\\
65,536 & 1.41 & 9.21 & 6.68 & 2.05\\ \bottomrule
  \end{tabular}  
\end{center}
\caption{Speedup obtained when using parallel linear recurrence compared to
  serial linear recurrence. We analyse this for several recent architectures:
  the Simple Recurrent Unit from \cite{lei2017}, the Quasi-RNN from
  \cite{bradbury2017quasi}, with a convolutional filter of size 2 and 10, and
  the GILR-LSTM which we introduce in section \ref{sec:ls-rnns}. To give a fair
  model of a typical use-case maximising the memory usage of the GPU, we set
  batch size \(b\) such that \(Tb = 65536\) for a sequence length \(T\).}
\end{figure}

Finally, we compare the throughput of the linear surrogate LSTM to the current
fastest available implementation of the LSTM, the CuDNN-LSTM. Figure
\ref{fig:tp_perf} shows that our GILR-LSTM has higher throughput for most of the
low-batch-size regime, with a 30x speedup at a sequence length of 65,536. Recent
work \cite{keskar2017large} has suggested that small mini-batches may aid
generalization by encouraging the discovery of shallow minima. The GILR-LSTM
allows us to achieve high throughput for all choices of minibatch size and
sequence length.
